---
title: To Settle a Score?
author: AR Domingo
date: '2021-04-14'
slug: to-settle-a-score
categories: 
- Politics
tags: 
- Events
---



<center>
<h4>It took one score, but could it really be true this time?</h4>
</center>


It's been a whole score since this war began, and there are now some fighting in it who weren't alive when it did.  It will have scarred **two** generations by the time it is finally over.  And that day might actually be within reach.<br>

Of course, as with any conflict, there are merits to the effort, and then there are failures.  And we have certainly had our share of each as a nation.  But an inter-generational war is a failure of imagination - after two decades, some progress must have been made to allow for some other means.  But to put this into perspective, there are about 80 million people aged 19 and younger in America.  That means that an **entire 25% of the country had not been born when this war started.**



```{r setup, message=FALSE, warning=FALSE, cache=TRUE,echo=FALSE}

library(tidyverse)
library(readxl)
library(lubridate)


afg <- read_xlsx("X:/OneDrive/02. Data/Military History/US Afghanistan War Deaths by Month.xlsx")


afg_long <- pivot_longer(afg, cols = JAN:DEC, names_to = "Month", values_to = "Deaths")



Date <- paste(afg_long$Year, afg_long$Month, rep("01", length(afg_long)), sep = "-")
Month<-as_date(Date)


afghan_war_table <- data.frame(Month, afg_long)
#colnames(afghan_war_table)[1] <- Date

colnames(afghan_war_table)[1] <- "Date"
colnames(afghan_war_table)[3] <- "Month"


levels <- c("DEC", "NOV", "OCT", "SEP", "AUG", "JUL", "JUN", "MAY", "APR", "MAR", "FEB", "JAN")     
afghan_war_table$Month <- factor(afghan_war_table$Month, levels = levels)


```




<center>

```{r vis1, echo=FALSE, message=FALSE, warning=FALSE}

library(tidyverse)
library(ggthemes)

afghan_vis <- ggplot(afghan_war_table,aes(x=Year,y=Month, fill=Deaths)) +
  geom_tile(color = "white", width=0.9, height=0.9) +
  geom_label(label = afghan_war_table$Deaths, color = "white") +
  theme_hc() +
  scale_fill_gradient(low="blue", high="red") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "US Deaths by Month - Afghanistan War (2001 - 2021?)", caption = "Source: Wikipedia") +
  theme(plot.title = element_text(hjust = 0.5))

afghan_vis


```
</center>


<center>
<h4>All's well that...well, ends.</h4>
</center>

<<<<<<< HEAD
President Biden has committed to ending this conflict at long last...but we've heard this line before.  So time will tell if he can succeed where his three predecessors failed.
=======
<<<<<<< HEAD
President Biden has committed to ending this conflict at long last...but we've heard this line before.  So time will tell if he can succeed where his three predecessors failed.
=======
President Biden has committed to ending this conflict at long last...but we've heard this line before.  So time will tell if he can succeed where his four predecessors failed.
>>>>>>> 86d21a78eecea5e48326b8c4167ac9b96cb7e306
>>>>>>> 96f51d50bf26585fe6d7872a2b39e12526fbcfcc

Admittedly, our "nation-building" efforts may not have that greatest success rate, at least in the way that you should consider success.  At the same time, I think we too often forget about where it has succeeded, and at times exceedingly.  Perhaps the beacon of triumph is that of South Korea.  The US isolated it from Soviet and Chinese influence as much as possible, and went about the years long effort to help build not just a nation, but cultivate a thriving free-market democracy.  

So maybe it's true that America's history in this respect is not always the exemplar, but she certainly has some moments.


