---
title: "MammalSlides"
author: "AR Domingo"
date: "4/25/2021"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


<center>
## Some of the smallest are the biggest
</center>

When it comes to shear number of differentiated species, the rodents take the cake and just about everything else.  

```{r rodents, echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.align='center'}

library(tidyverse)
library(waffle)
library(RColorBrewer)
library(wesanderson)
library(hrbrthemes)
library(formattable)

#Load data set
mdd <- read.csv("X:/OneDrive/02. Data/Environmental Data/Mammal Diversity Database/MDD/MDD/MDD_v1.4_6533species.csv")

#Create table
pivot_table_set <- 
  mdd %>% 
    group_by(order) %>%
    summarize(n())

#Rename columns
colnames(pivot_table_set) <- c("Order", "Number of Species")

#Arrange descending and add percent calculation
pivot_table_set <- 
  pivot_table_set %>% 
    arrange(desc(`Number of Species`)) %>% 
    mutate(Percent = `Number of Species`/sum(`Number of Species`))

#Create formattable object
order_table <- 
  formattable(pivot_table_set,
            align = c("c", rep("r", NCOL(pivot_table_set) -1)),
            list(
              `Number of Species` = color_bar("lightblue"),
              Percent = percent,
              Percent = color_tile("transparent", "skyblue")
            ))



#ALL MAMMALS
#Set main waffle dataframe
mdd %>%
  count(order) %>%
  ggplot(aes(fill = order, values = n)) +
  expand_limits(x=c(0,0), y=c(0,0)) +
  coord_equal() +
  labs(fill = NULL, colour = NULL) +
  theme_ipsum_rc(grid="") +
  theme_enhance_waffle() -> waf

#Create color set for all
colorset_all <- 
  c("#7FC97F", "#BEAED4", "#FDC086", "#7FFF00", "#BF5B17",
    "#666666", "#1B9E77", "#D95F02", "#7570B3", "#E7298A",
    "#66A61E", "#E6AB02", "#A6761D", "#A6CEE3", "#1F78B4",
    "#B3E2CD", "#FDDAEC", "#E41A1C", "#80B1D3", "#B3DE69",
    "#F781BF", "#1A1AFF", "#FB8072", "#A65628", "#CCEBC5",
    "#999999", "#F0027F")

#MAMMALS
mammalia_waffle <- 
  waf +
    geom_waffle(n_rows = 40, size = .5, colour = "white", flip = FALSE) +
    scale_fill_manual(values = colorset_all, length(order)) +
    theme(legend.title = element_blank()) +
    theme(legend.position="bottom")+
    theme(legend.text=element_text(size=7)) +
    labs(title = "MAMMALIA", caption = "Source: ASM Mammal Diversity Database") +
    theme(plot.title = element_text(hjust = 0.5))





#RODENTIA
colorset_rodentia <- 
  c(alpha("#7FC97F", .25), alpha("#BEAED4", .25), alpha("#FDC086", .25), alpha("#7FFF00", .25), alpha("#BF5B17", .25), 
    alpha("#666666", .25), alpha("#1B9E77", .25), alpha("#D95F02", .25), alpha("#7570B3", .25), alpha("#E7298A", .25),
    alpha("#66A61E", .25), alpha("#E6AB02", .25), alpha("#A6761D", .25), alpha("#A6CEE3", .25), alpha("#1F78B4", .25),
    alpha("#B3E2CD", .25), alpha("#FDDAEC", .25), alpha("#E41A1C", .25), alpha("#80B1D3", .25), alpha("#B3DE69", .25),
    alpha("#F781BF", .25), alpha("#1A1AFF", .25), alpha("#FB8072", .25), "#A65628",  alpha("#CCEBC5", .25),
    alpha("#999999", .25), alpha("#F0027F", .25))


rodentia_waffle <- 
  waf +
  geom_waffle(n_rows = 40, size = .5, colour = "white", flip = FALSE) +
  scale_fill_manual(values = colorset_rodentia, length(order)) +
  theme(legend.title = element_blank()) +
  theme(legend.position="bottom")+
  theme(legend.text=element_text(size=7)) +
  labs(title = "THE RODENTS", caption = "Source: ASM Mammal Diversity Database") +
  theme(plot.title = element_text(hjust = 0.5))

rodentia_waffle


```


<center>
## For the only flying ones, there's quite a variety
</center>


Chiroptera, the virtually omnipresent bat, gained more infamy following the events of 2020, but come in the second largest variety of all mammals - and are the only mammals that can fly.  


```{r bats, echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.align='center'}

#CHIROPTERA
colorset_chiroptera <- 
  c(alpha("#7FC97F", .25), alpha("#BEAED4", .25), alpha("#FDC086", .25), "#7FFF00", alpha("#BF5B17", .25), 
    alpha("#666666", .25), alpha("#1B9E77", .25), alpha("#D95F02", .25), alpha("#7570B3", .25), alpha("#E7298A", .25),
    alpha("#66A61E", .25), alpha("#E6AB02", .25), alpha("#A6761D", .25), alpha("#A6CEE3", .25), alpha("#1F78B4", .25),
    alpha("#B3E2CD", .25), alpha("#FDDAEC", .25), alpha("#E41A1C", .25), alpha("#80B1D3", .25), alpha("#B3DE69", .25),
    alpha("#F781BF", .25), alpha("#1A1AFF", .25), alpha("#FB8072", .25), alpha("#A65628", .25),  alpha("#CCEBC5", .25),
    alpha("#999999", .25), alpha("#F0027F", .25))


chiroptera_waffle <- 
  waf +
  geom_waffle(n_rows = 40, size = .5, colour = "white", flip = FALSE) +
  scale_fill_manual(values = colorset_chiroptera, length(order)) +
  theme(legend.title = element_blank()) +
  theme(legend.position="bottom") +
  theme(legend.text=element_text(size=7)) +
  labs(title = "THE BATS", caption = "Source: ASM Mammal Diversity Database") +
  theme(plot.title = element_text(hjust = 0.5))

chiroptera_waffle



```
