---
title: "COVID-19 County Severity Dashboard - CALIFORNIA"
output: 
  flexdashboard::flex_dashboard:
    theme: cerulean
    orientation: columns
    vertical_layout: fill
---



```{r  datadownload, include=FALSE, echo=FALSE, message=FALSE, cache=TRUE}


library(here)
library(bookdown)
library(flexdashboard)
library(tidyverse)
library(scales)
library(data.table)
library(DT)
library(RColorBrewer)
library(lubridate)
library(ggthemes)
library(plotly)
library(htmlwidgets)
library(sf)
library(leaflet)
library(maps)
library(htmltools)
library(viridis)
library(rgdal)
library(utils)
library(readxl)
library(tigris)


here("Covid Severity Dashboard")

#setwd("C:/Users/andre/OneDrive/01. Projects/CoViD-19")

#setwd(file.path("X:", "OneDrive", "01. Projects", "CoViD-19"))
here()

#Download current data
NYT_Cnty <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv", header = TRUE)
# NYT_State <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv", header = TRUE)


####################################################
#GET ALL CA ZIP CODES
#CA Zip Codes
#CAZ <- "X:/OneDrive/02. Data/Geographic Data/US Zip Code Shapefiles/CA/acs2019_5yr_B01003_86000US91916.zip"

options(tigris_use_cache = TRUE)

CAZ <- counties(state = "CA")


CA_Zipcodes_File <- unzip(CAZ)
#zipmap <- readOGR(dsn = path.expand("CA_Shapes/CA_Counties"), layer = "CA_Counties_TIGER2016" )


CA_Zipcodes <- st_read(CA_Zipcodes_File[3]) %>% 
  st_transform("+proj=longlat +datum=WGS84")


#Set up map
download.file("https://data.ca.gov/dataset/e212e397-1277-4df3-8c22-40721b095f33/resource/b0007416-a325-4777-9295-368ea6b710e6/download/ca-county-boundaries.zip",
              "CA_Shapes.zip")

CA1 <- unzip("CA_Shapes.zip")
#zipmap <- readOGR(dsn = path.expand("CA_Shapes/CA_Counties"), layer = "CA_Counties_TIGER2016" )


CA_areas1 <- st_read(CA1[6]) %>% 
  st_transform("+proj=longlat +datum=WGS84")


```



```{r setup, include=FALSE, echo=FALSE, message=FALSE}


#Filter for CA and convert dates
CA_Counties <- filter(NYT_Cnty, state == "California")
CA_Counties$date <- as.Date(CA_Counties$date)


#Process county by county data

#CA_County_Max <- as.data.table(CA_Counties)

CA_County_Max <- as.data.table(CA_Counties)
CA_County_Max <- na.omit(CA_County_Max)
#Find maximum for each county
CA_CaseMax <- CA_County_Max[CA_County_Max[, .I[which.max(cases)], by = county]$V1]

# CA_pops <- read.csv("C:/Users/andre/OneDrive/01. Projects/CoViD-19/CA County Population.csv", header = TRUE)

CA_pops <- read.csv("https://raw.githubusercontent.com/ardomingo/COVID-County-Severity-CA/main/CA%20County%20Population.csv", header = TRUE)

#Upload data and merge with NYT data
# CA_pops <- read.csv("C:/Users/andre/OneDrive/01. Projects/CoViD-19/CA County Population.csv", header = TRUE)
CA_merged <- merge(CA_CaseMax, CA_pops, by = "county")

#Add Rate/1000 calculation
CA_calc <- mutate(CA_merged, case_rate1k = round(1000 * (cases/population), digits = 2), 
                  death_rate1k = round(1000 * (deaths/population), digits = 2))


#Calculate recent hotspots
hotspot_set1 <- filter(CA_Counties, date == today()-days(1))
hotspot_set14 <- filter(CA_Counties, date == today()-days(15))


hotspot_comp <- merge(hotspot_set1, hotspot_set14, by = "county")
hotspot_comp1 <- merge(hotspot_comp, CA_pops, by = "county")

hotspot_comp_calc <- mutate(hotspot_comp1, caseslast14 = (cases.x - cases.y)/population *1000, 
                            deathslast14 = (deaths.x - deaths.y)/population*1000, 
                            case_perc = round(-((cases.y/cases.x)-1)*100, 2),
                            death_perc = round(-((deaths.y/deaths.x)-1)*100, 2))

hotspot_comp_calc$case_perc[is.na(hotspot_comp_calc$case_perc)] <- 0
hotspot_comp_calc$death_perc[is.na(hotspot_comp_calc$death_perc)] <- 0
#####


hotspot_comp_tots14 <- mutate(hotspot_comp1, caseslast14 = (cases.x - cases.y)/14, 
                             deathslast14 = (deaths.x - deaths.y)/14)






##################################

# CA_areas1 <- st_read("C:/Users/andre/OneDrive/01. Projects/CoViD-19/CA_Counties/CA_Counties_TIGER2016.shp") %>%
#   st_transform("+proj=longlat +datum=WGS84")
#########################################################


#CALCULATE HOT SPOTS LAST 14 DAYS

#Find +/-
hotspot_map_data <- merge(hotspot_comp_calc, hotspot_comp_tots14, by = "county")


hotspot_map_set <- merge(CA_areas1, hotspot_map_data, by.x = "NAME", by.y = "county")

pal <- colorRampPalette(c("blue", "purple"))

bar_colors <- colorRampPalette(viridis(length(CA_merged$county)))


#Setup City of Riverside Map

riv_zips <- read_xlsx("C:/Users/andre/OneDrive/01. Projects/CoViD-19/RiversideCityZips.xlsx")

RivCity <- riv_zips %>% 
  filter(ZIPCODE >= 92501 & ZIPCODE <= 92509)

RivCity <- RivCity[, 1:3] %>% 
  mutate(Change = round(100*(`New Cases` / `Total Cases **`), digits = 2))

RVSDZip <- unzip("C:/Users/andre/OneDrive/01. Projects/CoViD-19/Zip_Codes.zip")



RVSD <- st_read(RVSDZip[5]) %>% 
  st_transform("+proj=longlat +datum=WGS84")
riv_map_set <- merge(RVSD, RivCity, by.x = "ZIPCODE", by.y = "ZIPCODE")


```

# LAST 14 DAYS

Column {data-width=500}
-----------------------------------------------------------------------

### Percent Case Increase by County

```{r map, out.width= '100%'}


palC <- colorNumeric(
  palette = "Reds",
  domain = hotspot_map_set$case_perc)

map_popupsC <- paste(
  "County: ", hotspot_map_set$NAME, "|",
  "% Change: ", hotspot_map_set$case_perc, "", sep = " ")

map <-
leaflet(hotspot_map_set) %>%
  setView(-120, 37.7, 6) %>%
  addPolygons(stroke = TRUE, weight = .25, fillOpacity = .4,
              smoothFactor = 0.5, color = ~palC(case_perc),
              label = map_popupsC,
              highlightOptions = 
                highlightOptions(color = "white", weight = 1.75, opacity = 1,
                bringToFront = TRUE),
              labelOptions = labelOptions(style = list(
                "color" = "darkred",
                "font-family" = "serif",
                "font-variant" = "small-caps",
                "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
                "font-size" = "14px",
                "border-color" = "rgba(0,0,0,0.5)"))) %>% 
  addProviderTiles("CartoDB.Positron")

map %>% 
  config(displayModeBar = FALSE)
```

### Sortable Data Table

```{r table, out.width='100'}

#Add data table
table_set <- hotspot_map_data[, c(1,5,6,12,15,16)]
colnames = c("County", "Cases", "Deaths", "Population", "Cases % Change", "Deaths % Change")  #Case Rate per 1000")

datatable(table_set, colnames = c("County", "Total Cases", "Total Deaths", "Population", "Cases 14-Day % Change", "Deaths 14-Day % Change"), 
          rownames = FALSE, options = list(order = list(list(4, "desc")), pageLength = 2,
                                           rowCallback = JS("function(r,d) {$(r).attr('height', '75px')}")))

```




Column {data-width=500}
-----------------------------------------------------------------------

### Cases - Last 14 Days

```{r bar14, message=FALSE, warning=FALSE, out.width='100'}
#Plot bar chart for cases last 14 days
bar_cases_last14 <-
  ggplot(data = hotspot_comp_tots14, aes(x = reorder(county, -caseslast14), y = caseslast14, 
                                         fill = county, group = 1, 
                                         text = paste("Cases: ", caseslast14, "\n",
                                                      "County: ", county, sep = ""))) +
  geom_bar(stat = "identity") +
  labs(title = "Average Daily Cases - Last 14 Days", x = "County") +
  scale_y_continuous(name = "Confirmed Cases", label = comma) +
  theme_classic() +
  theme(axis.text.x=element_text(angle=90,hjust=0,vjust = .4)) +
  theme(legend.position = "none") +
  scale_fill_manual(values = bar_colors(58))


ggplotly(bar_cases_last14, tooltip = c("County", "text"))

```


# CASES

Column {data-width=500}
-----------------------------------------------------------------------

### Time Series

```{r line case, echo=FALSE, message=FALSE, warning=FALSE, out.width='100'}

#Linechart cases
line_cases <-
  ggplot() +
  geom_line(data = CA_Counties, aes(x = date, y = cases, color = county, group = county,
                                    text = paste("Date: ", date, "\n",
                                                 "Cases: ", cases, "\n",
                                                 "County: ", county,
                                                 sep = "")), size = .5, alpha = .5) +
  scale_y_continuous(name = "Confirmed Cases", labels = comma)+
  scale_x_date(name = "Month") + 
  labs(title = "Confirmed Cases", subtitle = "CA Counties", caption = "Fig. 1") +
  theme_classic() +
  theme(axis.text.x=element_text(angle=90,hjust=0,vjust = .4)) +
  theme(legend.position = "none") +
  scale_color_manual(values = bar_colors(length(unique(CA_Counties$county))))

ggplotly(line_cases, tooltip = "text")

```







Column {data-width=500}
-----------------------------------------------------------------------

### Total Cases per 100,000 by County

```{r barchart caserate1k, echo=FALSE, message=FALSE, warning=FALSE, out.width='100'}

###Plot barchart of rate data###

#Normalize data by county population

#Upload data and merge with NYT data
CA_pops <- read.csv("C:/Users/andre/OneDrive/01. Projects/CoViD-19/CA County Population.csv", header = TRUE)
CA_merged <- merge(CA_CaseMax, CA_pops, by = "county")


#Add Rate/1000 calculation
CA_calc <- mutate(CA_merged, case_rate1k = round(100000 * (cases/population), digits = 2), 
                  death_rate1k = round(1000 * (deaths/population), digits = 2))


#Plot bar chart cases
bar_cases1k <- 
ggplot(data = CA_calc, aes(x = reorder(county, -case_rate1k), y = case_rate1k, 
                           fill = county, group = 1, 
                           text = paste("Cases: ", case_rate1k, "\n",
                                        "County: ", county, sep = ""))) +
  geom_bar(stat = "identity") +
  labs(title = "Confirmed Cases per 100,000", subtitle = "CA Counties", x = "County") +
  scale_y_continuous(name = "Total Confirmed Cases", label = comma) +
  theme_classic() +
  theme(axis.text.x=element_text(angle=90,hjust=0,vjust = .4)) +
  theme(legend.position = "none") +
  geom_hline(yintercept = median(CA_calc$case_rate1k), color="blue", alpha = .25) +
  scale_fill_manual(values =bar_colors(length(unique(CA_Counties$county))))

ggplotly(bar_cases1k, tooltip = c("County", "text"))


```






# DEATHS

Column {data-width=500}
-----------------------------------------------------------------------

### Time Series

```{r line death, echo=FALSE, message=FALSE, warning=FALSE, out.width='100'}


#Linechart deaths

line_deaths <-
  ggplot() +
  geom_line(data = CA_Counties, aes(x = date, y = deaths, color = county, group = county,
                                    text = paste("Date: ", date, "\n",
                                                 "Deaths: ", deaths, "\n",
                                                 "County: ", county,
                                                 sep = "")), size = .5, alpha = .5) +
  scale_y_continuous(name = "Total Confirmed Deaths", labels = comma)+
  scale_x_date(name = "Month") + 
  labs(title = "Confirmed Deaths", subtitle = "CA Counties", caption = "Fig. 1") +
  theme_classic() +
  theme(axis.text.x=element_text(angle=90,hjust=0,vjust = .4)) +
  theme(legend.position = "none")+
  scale_color_manual(values = bar_colors(length(unique(CA_Counties$county))))

ggplotly(line_deaths, tooltip = "text")


```



Column {data-width=500}
-----------------------------------------------------------------------

### Total Deaths per 1,000 by County

```{r barchart deathrate1k, echo=FALSE, message=FALSE, warning=FALSE, out.width='100'}

#Plot bar chart for deaths per 1000
bar_deaths1k <-
ggplot(data = CA_calc, aes(x = reorder(county, -death_rate1k), y = death_rate1k, fill = county, group = 1,
                           text = paste("Deaths: ", death_rate1k, "\n",
                                        "County: ", county, sep = ""))) +
  geom_bar(stat = "identity") +
  labs(title = "Confirmed Deaths per 1000", subtitle = "CA Counties", x = "County", caption = "Fig. 6") +
  scale_y_continuous(name = "Confirmed Cases", label = comma) +
  theme_classic() +
  theme(axis.text.x=element_text(angle=90,hjust=0,vjust = .4)) +
  theme(legend.position = "none") +
  geom_hline(yintercept = median(CA_calc$death_rate1k), color="blue", alpha = .25)+
  scale_fill_manual(values =bar_colors(length(unique(CA_Counties$county))))

ggplotly(bar_deaths1k, tooltip = c("County", "text"))


```





CITY OF RIVERSIDE {.tabset}
===================================== 


## City of Riverside by Zipcode 

<font size="2"><em>*Most recent data available are as of 04/02/2021.</em></font><br><br>

```{r rivcity, echo=FALSE, message=FALSE, warning=FALSE, out.width='100%'}



datatable(RivCity, rownames = FALSE, colnames = c("Zip Code", "Total Cases", "New Cases", "% Change"))


```


## Riverside Map 


```{r riv_map, echo=FALSE, message=FALSE, warning=FALSE, out.width='100%'}


map_popups_rivcity <- paste(
  "Zipcode: ", riv_map_set$ZIPCODE, "|",
  "% Change: ", riv_map_set$Change, "", sep = " ")

palRivCity <- colorNumeric(
  palette = "Reds",
  domain = riv_map_set$Change)


rivcitymap <-
  leaflet(riv_map_set) %>%
  setView(-117.39535758810685, 33.93643526692571, 11) %>% 
  addPolygons(stroke = TRUE, weight = .25, fillOpacity = .4,
              smoothFactor = 0.5, color = ~palRivCity(Change),
              label = map_popups_rivcity,
              highlightOptions = 
                highlightOptions(color = "white", weight = 1.75, opacity = 1,
                                 bringToFront = TRUE),
              labelOptions = labelOptions(style = list(
                "color" = "darkred",
                "font-family" = "serif",
                "font-variant" = "small-caps",
                "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
                "font-size" = "14px",
                "border-color" = "rgba(0,0,0,0.5)"))) %>% 
  addProviderTiles("CartoDB.Positron")

rivcitymap %>% 
  config(displayModeBar = FALSE)

```



CITY OF CORONA {.tabset}
===================================== 


## City of Corona by Zipcode 


```{r corona_city_setup, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
#Setup City of Corona Map


corona_zips <- read_xlsx("C:/Users/andre/OneDrive/01. Projects/CoViD-19/RiversideCityZips2.xlsx")

CoronaCity <- corona_zips %>% 
  filter(ZIPCODE >= 92879 & ZIPCODE <= 92883)

CoronaCity <- CoronaCity[, 1:3] %>% 
  mutate(Change = round(100*(`New Cases` / `Total Cases **`), digits = 2))


CA_Zipcodes <- st_read(CA_Zipcodes_File[3]) %>% 
  st_transform("+proj=longlat +datum=WGS84")

corona_map_set <- merge(CA_Zipcodes, CoronaCity, by.x = "name", by.y = "ZIPCODE")
```


```{r corona_city, echo=FALSE, message=FALSE, warning=FALSE, out.width='100%'}


datatable(CoronaCity, rownames = FALSE, colnames = c("Zip Code", "Total Cases", "New Cases", "% Change"))

```



## Corona Map 


```{r corona_map, echo=FALSE, message=FALSE, warning=FALSE, out.width='100%'}


map_popups_coronacity <- paste(
  "Zipcode: ", corona_map_set$ZIPCODE, "|",
  "% Change: ", corona_map_set$Change, "", sep = " ")

palCoronaCity <- colorNumeric(
  palette = "Reds",
  domain = corona_map_set$Change)



coronacitymap <-
  leaflet(corona_map_set) %>%
  setView(-117.56173120190176, 33.869114757649356, 11) %>% 
  addPolygons(stroke = TRUE, weight = .25, fillOpacity = .4,
              smoothFactor = 0.5, color = ~palCoronaCity(Change),
              label = map_popups_coronacity,
              highlightOptions = 
                highlightOptions(color = "white", weight = 1.75, opacity = 1,
                                 bringToFront = TRUE),
              labelOptions = labelOptions(style = list(
                "color" = "darkred",
                "font-family" = "serif",
                "font-variant" = "small-caps",
                "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
                "font-size" = "14px",
                "border-color" = "rgba(0,0,0,0.5)"))) %>% 
  addProviderTiles("CartoDB.Positron")

coronacitymap %>% 
  config(displayModeBar = FALSE)
```



CITY OF MORENO VALLEY {.tabset}
===================================== 


## City of Moreno Valley by Zipcode 


```{r moval_city_setup, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
#Setup City of Corona Map


moval_zips <- read_xlsx("C:/Users/andre/OneDrive/01. Projects/CoViD-19/RiversideCityZips2.xlsx")

MoValCity <- moval_zips %>% 
  filter(ZIPCODE >= 92551 & ZIPCODE <= 92557)

MoValCity <- MoValCity[, 1:3] %>% 
  mutate(Change = round(100*(`New Cases` / `Total Cases **`), digits = 2))


moval_map_set <- merge(CA_Zipcodes, MoValCity, by.x = "name", by.y = "ZIPCODE")
```


```{r moval_city, echo=FALSE, message=FALSE, warning=FALSE, out.width='100%'}


datatable(MoValCity, rownames = FALSE, colnames = c("Zip Code", "Total Cases", "New Cases", "% Change"))

```



## Moreno Valley Map 


```{r moval_map, echo=FALSE, message=FALSE, warning=FALSE, out.width='100%'}


map_popups_movalcity <- paste(
  "Zipcode: ", moval_map_set$ZIPCODE, "|",
  "% Change: ", moval_map_set$Change, "", sep = " ")

palMoValCity <- colorNumeric(
  palette = "Reds",
  domain = moval_map_set$Change)


movalcitymap <-
  leaflet(moval_map_set) %>%
  setView(-117.22584072103172, 33.916212657818946, 11) %>% 
  addPolygons(stroke = TRUE, weight = .25, fillOpacity = .4,
              smoothFactor = 0.5, color = ~palMoValCity(Change),
              label = map_popups_movalcity,
              highlightOptions = 
                highlightOptions(color = "white", weight = 1.75, opacity = 1,
                                 bringToFront = TRUE),
              labelOptions = labelOptions(style = list(
                "color" = "darkred",
                "font-family" = "serif",
                "font-variant" = "small-caps",
                "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
                "font-size" = "14px",
                "border-color" = "rgba(0,0,0,0.5)"))) %>% 
  addProviderTiles("CartoDB.Positron")

movalcitymap %>% 
  config(displayModeBar = FALSE)
```

