---
title: No, it really might be a GOAT
author: AR Domingo
date: '2021-04-10'
slug: no-it-really-might-be-a-goat
categories:
  - Sports
tags:
  - Football
  - NFL
  - formattable
---

<center>
## Does it Quack?
</center>


```{r setup, message=FALSE, warning=FALSE, include=FALSE, cache=TRUE}


library(tidyverse)
library(DT)


#Import data

#Get yearly team passing stats
teampass_2015 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataTeamPassing2015.csv") %>% 
  summarize_if(is.numeric, mean)
teampass_2015$Season <- 2015


teampass_2016 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataTeamPassing2016.csv") %>% 
  summarize_if(is.numeric, mean)
teampass_2016$Season <- 2016


teampass_2017 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataTeamPassing2017.csv") %>% 
  summarize_if(is.numeric, mean)
teampass_2017$Season <- 2017


teampass_2018 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataTeamPassing2018.csv") %>% 
  summarize_if(is.numeric, mean)
teampass_2018$Season <- 2018


teampass_2019 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataTeamPassing2019.csv") %>% 
  summarize_if(is.numeric, mean)
teampass_2019$Season <- 2019

teampass_2020 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataTeamPassing2020.csv") %>% 
  summarize_if(is.numeric, mean)
teampass_2020$Season <- 2020


#Combine 15-19 into table
league_avgs <- rbind(teampass_2015, teampass_2016, teampass_2017, teampass_2018, teampass_2019)

#Prep df's to match to bind
league1 <- league_avgs[ , c(1:2, 5:7)] %>% 
    rename(Yds = Pass.Yds,
           Int = INT)



#Player stats
passing_2015 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataPassing2015.csv")
passing_2015$Year <- 2015

passing_2016 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataPassing2016.csv")
passing_2016$Year <- 2016

passing_2017 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataPassing2017.csv")
passing_2017$Year <- 2017

passing_2018 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataPassing2018.csv")
passing_2018$Year <- 2018

passing_2019 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataPassing2019.csv")
passing_2019$Year <- 2019

passing_2020 <- read.csv("X:/OneDrive/02. Data/NFL Data/NFl DataPassing2020.csv")
passing_2020$Year <- 2020

#Combine sets and reorder
passing_combined_15to19 <- rbind(passing_2015, passing_2016, passing_2017, passing_2018, passing_2019)
passing_combined_15to19 <- passing_combined_15to19[ , c(1, 4:5, 7, 9, 11, 17)]





#Import Brady Career Stats
tom_brady <- read.csv("X:/OneDrive/02. Data/NFL Data/TomBradyCareer.csv")
tom_brady <- tom_brady[1:20, c(1:6, 8, 10, 12, 14)]

tom_brady_to14 <- tom_brady %>% 
  filter(Year < 2015)

tom_brady_15on <- tom_brady %>% 
  filter(Year >= 2015)

#Get Totals for each era
brady_0014 <- 
  tom_brady_to14 %>% summarise_if(is.numeric, sum)

brady_1519 <- 
  tom_brady_15on %>% summarise_if(is.numeric, sum)

career_splits <- 
  rbind(brady_0014, brady_1519)


#Get Averages for each era
brady_0014_avg <- 
  tom_brady_to14 %>% summarise_if(is.numeric, mean)

brady_1519_avg <- 
  tom_brady_15on %>% summarise_if(is.numeric, mean)

career_splits_avg <- 
  rbind(brady_0014_avg, brady_1519_avg)

rownames(career_splits_avg) <- c("Early", "Late")



brady_early <- brady_0014_avg[ , 3:7]
brady_early$Years <- "2000-2014"
brady_early$Party <- "BRADY Avg."

brady_late <- brady_1519_avg[ , 3:7]
brady_late$Years <- "2015-2019"
brady_late$Party <- "BRADY Avg."


#Prep df's to match to bind
league1 <- league_avgs[ , c(1:2, 5:7)] %>% 
    rename(Yds = Pass.Yds,
           Int = INT)

league1$Years <- c(2015:2019)
league1$Party <- "League Avg."



brady_late_insert <- brady_late[, 1:6] %>% 
  rename(Year = Years)
brady_late_insert$Player <- "T. BRADY"


brady2020 <- filter(passing_2020, grepl("Brady", Player, fixed = TRUE))
brady2020 <- brady2020[ , c(1, 4:5, 7, 9, 11, 17)]

#Create table with {DT} pkg
avgs_table_set <- rbind(league1, brady_early, brady_late)

avgs_table_set <- avgs_table_set[, c(7, 6, 1:5)]
avgs_table_set <- avgs_table_set[order(avgs_table_set$Party, desc(avgs_table_set$Year)), ]

#Add Brady 2020 stats for emphasis
percentiles <- rbind(passing_combined_15to19, brady_late_insert, brady2020)

percentiles <- 
percentiles %>% 
  mutate(AttPCT = ntile(Att, 100),
         CmpPCT = ntile(Cmp, 100),
         YdsPCT = ntile(Yds, 100),
         TDPCT = ntile(TD, 100),
         IntPCT = 100-ntile(Int, 100),
         Cmpl = Cmp / Att,
         CmplPCT = ntile(Cmpl, 100)
         )

#Reorder
percentiles <- percentiles[, c(1, 7, 2:3, 13, 4:6, 8:9, 14, 10:12)]


#######  BRADY vs The Players data

league_chart_PCTs <- percentiles[,c(1:2, 9:14)]
#league_chart_trim1 <- league_chart_trim[ , c(1, 3:8)]

 
league_chart_PCTs %>% 
group_by(Player, Year) %>% 
  summarise(Year, AttPCT, CmpPCT, CmplPCT, YdsPCT, TDPCT, IntPCT) 
  

brady_PCTs <- filter(league_chart_PCTs, grepl("Brady|BRADY", Player))
brady_PCTs <- brady_PCTs[order(brady_PCTs$Player, desc(brady_PCTs$Year)), ]



```



On the day of Super Bowl LV before the game got under way, I noted [this](https://repartee-per-se.netlify.app/post/if-it-walks-like-a-goat/) about Tom Brady.  This was, admittedly, based on not much more than a gut feeling I got about his performance over his career, as well as my perceptions of him throughout his time in the NFL.  But I needed to know - exactly how good had Brady been during that five years when he could have been "retired" and waiting to get voted into the HOF?  What did he do with himself instead of putting on the production of the triumphant Hall of Famer returning to play for one more ring?  What did he do actually playing football in those five years?


<center>
### Brady v. The League
</center>

If you look at the more vital passing statistics including attempts (Att), completions (Cmp), passing yards (Yds), touchdowns (TD), and interceptions (Int), it becomes pretty clear how Brady spent his time.  His averages for the five seasons between 2015-2019 outpaced what the NFL averaged in each of those seasons in 4 out of 5 of those categories.  Brady's averages over that span, against the league:


- Rank **3rd** in Attempts
- Rank **1st** in Completions
- Rank **1st** in Yards
- Rank **1st** in Touchdowns
- Rank **1st** in Interceptions

<br>


<center>
```{r brady v league, echo=FALSE, message=FALSE, warning=FALSE, fig.width=100}

options(digits = 5)  

library(tidyverse)
library(DT)
library(formattable)

  
row.names(avgs_table_set) <- NULL

formattable(avgs_table_set,
            list(
              Att = color_tile("transparent", "pink"),
              Cmp = color_tile("transparent", "lightgray"),
              Yds = color_tile("transparent", "lightblue"),
              TD = color_tile("transparent", "pink"),
              Int = color_tile("lightgray", "transparent"))) 

# datatable(avgs_table_set,           
#       rownames = FALSE,
#       options = list(
#             dom = 't',
#             order = list(list(2, 'dsc')),
#             columnDefs = list(list(className = 'dt-center', targets = 0:6)),
#       initComplete = JS("
#                          function(settings, json) {
#                          $(this.api().table().header()).css({
#                          'background-color': 'silver',
#                          'color': 'blue'
#                          });
#                          }"))) %>%
#             formatRound(columns=c(1:5), digits=2)


```
</center>

<center>
#### Brady v. The Players
</center>

So good old #12's performance over those years was essentially a clinic for the league, on average.  But how does he stack up against individual players?  By looking at how his stats rank, as a percentile, against all other QBs gives a pretty good idea that he was generally at least above average.^[By taking every QB in the NFL with at least 14 pass attempts per game and compiling their seasons stats, I then ranked them by percentile, with numbers closest to 100 being the league leaders in that stat category.  So this table provides how each of Brady's seasons (2015-2019; and 2020 added for context) **and** his average performance, each compare, in terms of his percentile rankings, against the seasons of every other QB's season totals during **every** season (excluding 2020).]  Indeed, most of his stats put his in at least the 60th percentile.  So while not always in a spot to win, place, or show, you would have been wise to grab him for your fantasy league. <br><br>


```{r brady v players, echo=FALSE, message=FALSE, warning=FALSE, fig.width=100}

library(tidyverse)
library(formattable)
library(DT)

row.names(brady_PCTs) <- NULL

formattable(brady_PCTs,
  list(
  AttPCT = color_bar("pink"),
  CmpPCT = color_bar("silver"),
  CmplPCT = color_bar("lightblue"),
  YdsPCT = color_bar("pink"),
  TDPCT = color_bar("silver"),
  IntPCT = color_bar("lightblue")))

```

He also won a couple of Super Bowls in that span for good measure...and then, with another career year this past season, with a different team, in a different conference, *he won another*.

<br><br>